Enhanced Material Point Method to face dynamic problems:\thanks{Funding: This study was funded by Agustín de Betancourt Foundation (grant number 262390106114).}

### 2020-03-20 main.tex
Cambios 14-3-2020
Cambios 20-3-2020

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
%\documentclass[smallextended]{svjour3}       % onecolumn (second format)
\documentclass[twocolumn]{svjour3}          % twocolumn
%
% flush right qed marks, e.g. at end of proof
\smartqed

@@

\title{Enhanced Material Point Method to face
  dynamic problems:\thanks{Funding: This
    study was funded by Agustín de Betancourt Foundation (grant number
    262390106114).}
}
\subtitle{Local-maximum entropy approximation and an explicit
  predictor-corrector scheme}

@@

\begin{algorithm}
  \label{alg:PCE-algorithm}
  \floatname{algorithm}{Algorithm}
  \renewcommand{\thealgorithm}{}
  \caption{Explicit Predictor-Corrector scheme}
  \begin{algorithmic}[1]
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% º
    \STATE \textbf{Explicit Newmark Predictor} ($\gamma = 0.5$):
    \begin{equation*}
      \vec{v}_I^{k+1} = \vec{v}_I^n + (1 - \gamma)\ \Delta t\ \vec{a}_I^k
      \quad , \quad \vec{a}_I^k = \frac{\vec{f}_{I}^{k}}{\tens{m}_I^k}      
    \end{equation*}
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Impose essential boundary conditions}:\\
    At the fixed boundary, set $\vec{v}_{I}^{k+1} = 0$. 
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    % \STATE \textbf{Discard the previous nodal values}.
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Update mass matrix}:
    \begin{equation*}
      \tens{m}_{I} = \sum_{p}^{N_p}N_{Ip}^{k+1}\ m_p,
    \end{equation*}
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Deformation tensor increment calculation}.
    \begin{align*}
      &\dot{\tens{\varepsilon}_{p}}^{k+1} = \frac{1}{2}\ \sum_{I}^{N_n}
        \Grad{N_{Ip}^{k+1}}\vec{v}_{I}^{k+1} +
        (\Grad{N_{Ip}^{k+1}}\vec{v}_{I}^{k+1})^T \\
      &\Delta \tens{\varepsilon}_{p}^{k+1} = \Delta t\ \dot{\tens{\varepsilon}_{p}}^{k+1}
    \end{align*}
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Update the density field}:
    \begin{equation*}
      \rho_p^{k+1} = \frac{\rho_p^k}{1 + \mathit{tra}\left[\Delta\tens{\varepsilon}_{p}^{k+1}\right]}.
    \end{equation*}
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Balance of forces calculation}:\\
    Calculate the total grid nodal force $\vec{f}_{I}^{k+1} =
    \vec{f}_{I}^{int,k+1} + \vec{f}_{I}^{ext,k+1}$ evaluating
    \eqref{eq:nodal_internal_forces} and \eqref{eq:nodal_external_forces} in the time step $k+1$. In the
    grid node $I$ is fixed in one of the spatial dimensions, set it to
    zero to make the grid nodal acceleration zero in that direction.\\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Explicit Newmark Corrector} ($\gamma = 0.5$):
    \begin{equation*}
      \vec{v}_{I}^{k+1} = \vec{v}_{I}^{k+1} + \gamma\ \Delta t\
      \vec{a}_I^{k+1} \quad , \quad  \vec{a}_I^{k+1} = \frac{\vec{f}_{I}^{k+1}}{\tens{m}_I^{k+1}}  
    \end{equation*}
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \STATE \textbf{Update particles position and velocity}:
    \begin{align*}
      &\vec{x}_p^{k+1} = \vec{x}_p^n + \Delta t\
        \sum^{N_n}_{I} N_{Ip}^k\ \vec{v}_{I}^{k} +
        \frac{1}{2}\Delta t^2\ \sum^{N_n}_{I}\frac{N_{Ip}^k\
        \vec{f}_{I}^{k}}{\tens{m}_I^k}\\
      &\vec{v}_p^{k+1} = \vec{v}_p^n + \Delta t\
        \sum^{N_n}_{I}\frac{N_{Ip}^k\
        \vec{f}_{I}^{k}}{\tens{m}_I^k}
    \end{align*}
  \end{algorithmic}
\end{algorithm}

@@

The previous lines describe the main algorithm without defining an
interpolation technique. A direct consequence of this, is that it is
possible to adopt a width range of them. The MPM has been
successfully applied to a wide range of problems due to its ability to
deal with large strain problems without mesh distorsion issues inherent to
mesh based methods like FE. However, in the simulations made with the
original MPM, there are numerical noises when particles crossing
the cell boundaries. Recently the local maximum-entropy (or local
\textit{max-ent}) shape function first introduced by Arroyo \& Ortiz
(2006)\cite{Arroyo2006} has been tested under the MPM framework by
Wobbes {\it et al.} (2020)\cite{Wobbes2020} where they prof that
simulations performed with the maxent basis functions show
considerably more accurate stress approximations for MPM. Although, in
\cite{Wobbes2020} authors does not deep in $\beta$ parameter
benefits. In this publication a parametric analysis of $\Beta$ will be
performed in order to reach a compromise value that minimize the
presence of wiggle without introducing an additional spurious behaviour.

The maximum-entropy estimate is defined as the type of statistical
inference, which is the least biased possible on the given information
\cite{Jaynes1957}. The basic idea of the shape functions based on such
an estimate is to interpret the shape function $N_I(\vec{x})$ as the
probability of $\vec{x}$ to obtain the value $\vec{x}_I$,  $I=1,
\dots, n$. Here $n$ is the number of nodes in the domain.
Taking Shannon's entropy as the starting point:
\begin{equation}
  \label{eq:Shannon-entropy}
  H(p_1(\vec{x}),\ldots,p_n(\vec{x})) = -\sum^{N_n}_{I=1}{p_I(\vec{x})\log p_I }
\end{equation}
where $p_I(\vec{x})$ is the probability, equivalent to the mentioned
shape function $N_I(\vec{x})$,  satisfying both the zeroth and
first-order consistency. The least-biased approximation scheme is given by
\begin{align*}
  \text{(LME)} \hspace{0.15cm} \text{Maximize} \hspace{0.15cm} &H(p) = -\sum_{I}^{N_n}{p_I(\vec{x})\log p_I } \nonumber \\
  \text{subject to} \hspace{0.15cm} &p_I \ge 0, \hspace{0.15cm} \text{I=1, ..., n} \nonumber \\
  &\sum_{I=1}^{N_n}{p_I} = 1 \nonumber \\
  &\sum_{I=1}^{N_n}{p_I \vec{x}_I} = \vec{x} \nonumber 
\end{align*}
The local max-ent approximation schemes (LME) as a Pareto set,
defined by \cite{Arroyo2006} is as follows 
\begin{align*}
  \text{(LME)}_{\beta} \hspace{0.15cm} \text{For fixed} \hspace{0.15cm}
  \vec{x} \hspace{0.15cm} \text{minimise}  \nonumber \hspace{0.15cm} &f_{\beta}(\vec{x}, p) = \beta H(\vec{x},p) - H(p) \\
  \text{subject to} \hspace{0.15cm} &p_I \ge 0, \hspace{0.2cm} \text{I=1, ..., n}  \\
                                                                   &\sum_{I=1}^{N_n}{p_I} = 1 \\
                                                                   &\sum_{I=1}^{N_n}{p_I \, \vec{x}_I} = \vec{x}
\end{align*}
for $\beta \in (0,\infty)$ is Pareto optimal. The unique solution of
the local max-ent problem (LME)$_\beta$ is: 
\begin{equation}
  \label{eq:LME-p}
p(\vec{x})=\frac{\exp\left[ -\beta \; |\vec{x}-\vec{x}_I|^2 +  \vec{\lambda}  \;  (\vec{x}-\vec{x}_I)         \right] } {Z(\vec{x},\vec{\lambda}^*(\vec{x}))}
\end{equation}
where
\begin{equation}
  \label{eq:LME-Z}
Z(\vec{x}, {\vec{\lambda}}) = \sum_{I=1}^{N_n}{ \exp \left[ -\beta \, | \vec{x}-\vec{x}_I|^2 + \vec{\lambda}  \,  (\vec{x}-\vec{x}_I)  \right]}
\end{equation}
being $\vec{\lambda}^*(\vec{x})$ the unique minimiser  for $\log Z(\vec{x}, \vec{\lambda})$

In order to obtain the first derivatives of the shape function, it is  also necessary to compute~$\nabla p^*_I$ 
\begin{equation}
  \label{eq:LME-grad-p}
\nabla p^*_I=p^*_I  \, \left(\nabla f^*_I-\sum_J^{N_n} p^*_J \, \nabla f^*_J\right)
\end{equation}
where
\begin{equation}
  \label{eq:LME-f}
f^*_I(\vec{x},  \vec{\lambda},\beta)=-\beta \, |\vec{x}-\vec{x}_I|^2 + \vec{\lambda}   \,  (\vec{x}-\vec{x}_I)
\end{equation}
Employing the chain rule, rearranging and considering $\beta$ as a constant, Arroyo and Ortiz~\cite{Arroyo2006} obtained the following expression:
\begin{eqnarray}
\nabla p^*_I &=& -p^*_I \,  (\tens{J}^*)^{-1} \,  (\vec{x} - \vec{x}_I) \label{eq26} 
\end{eqnarray}
%&&\mbox{
where \tens{J} is the Hessian matrix, defined by:
%} \nonumber \\&& \nonumber \\
\begin{eqnarray}
\tens{J}(\vec{x},  \vec{\lambda},\beta) &=& \frac{\partial \vec{r}}{\partial  \vec{\lambda}} \label{eq27} \\ 
%&& \nonumber \\
\vec{r}(\vec{x},\vec{\lambda},\beta) &\equiv& \partial_{\vec{\lambda}}{\log{ Z( \vec{x},\vec{\lambda}})} 
%\nonumber \\ &=&
  =\sum_I^{N_n} p_I(\vec{x},\vec{\lambda},\beta) \, (\vec{x} - \vec{x}_I) \label{eq28} 
\end{eqnarray}
Note that, the objective of the above procedure is to find the
$\vec{\lambda}$ which minimises  $\log Z(\vec{x}, \vec{\lambda})$. The
traditional way to obtain such a minimiser  is using Eq.~(\ref{eq27})
to calculate small increments of $\partial\vec{\lambda}$ in a
Newton-Raphson approach. 
Similar to alternative non-polynomial meshfree basis functions, the
LME approximation scheme requires more than $d+1$ nodes to determine
the values of the shape functions as well as their derivatives at any
point in the convex hull of the nodal set, where $d$ is the dimension
of the problem. Due to the FE-compatibility, the LME shape function is
degenerated to linear finite element shape function if $d+1$
neighbouring nodes are chosen as the support. The support size of LME
shape functions may be controlled by adjusting a dimensionless
parameter, $\gamma=\beta h^2$ (e.g. two dimensional example shown in
figure. \ref{fig:LME_MPM})\cite{Arroyo2006}.

@@

Since  $p_I$  is defined in the entire domain, in practice, the
function $\exp(-\beta \vec{r} )$ truncated  by  a given tolerance,
10$^{-6}$, for example,  would ensure a reasonable range of
neighbours, see \cite{Arroyo2006} for details. This tolerance defines
the limit values of the influence radius and is used thereafter to
find the neighbour nodes of a given integration point.
In this research and in \cite{Arroyo2006}, $\Beta$ is a scalar as the
influence area of the shape function is controlled by the Euclidean
norm, therefore the search area is geometrically a circle in 2D, or a
sphere in 3D. Building upon the idea of anisotropic shape functions,
\cite{Kochmann2019} introduced an enhanced version of the original
local max-ent scheme, which uses an anisotropic support to deal with
tensile inestability. Nonetheless this is out of the scope of the
present document but will be incorporated in future research.

% \subsection{Error analysis MPM vs uGIMP vs MPM-\textit{max-ent}}
% \label{sec:study-error}

% For the convergence analysis, the root-mean-square (RMS) error in the
% displacement is computed. RMS error is defined as follows

% \begin{equation}
%   \label{eq:RMS}
%   RMS = \sqrt{\frac{1}{N} \sum^{N}_p \left( \phi_p - \hat{\phi}_p \right)^2},
% \end{equation}

% where $\phi_p$ and $\hat{\phi}_p$ are respectively the analytical and
% numerical solutions evaluated in the final time step in the position
% of each particle.

@@

\begin{figure}\sidecaption
  \centering
  \resizebox{\hsize}{!}{
    \includegraphics[width=\textwidth]{./Figures/FE_vs_PCE_CFL_05}
  }
  \caption{Velocity evolution at the point in the bar left side.}
  \label{fig:Dyka-PCE-FE}
\end{figure}

@@

\begin{figure}
  \centering
  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
  \subfloat[t = 0 seconds.]{
    \includegraphics[width=0.8\columnwidth]{Figures/Block_LME3_PCE_t0}
    \label{fig:Block-LME3-PCE-t0}
  }
  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
  \qquad
  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
  \subfloat[t = 5 seconds.]{
    \includegraphics[width=0.8\columnwidth]{Figures/Block_LME3_PCE_t025}
    \label{fig:Block-LME3-PCE-t025}
  }
  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \qquad
  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
  \subfloat[t = 20 seconds]{
    \includegraphics[width=0.8\columnwidth]{Figures/Block_LME3_PCE_t1}
    \label{fig:Block-LME3-PCE-t1}
  }
  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
  \caption{Vertical normal stress and position of material points
    during the loading process for a soft soil ($E = 5\ MPa$, $\rho_0
    = 6\cdot 10^3\ kg/m^3$). Numerical parameters considered for the
    simulation are : Local \textit{max-ent} shape function $\gamma =3$
    and explicit PC scheme with CFL 0.1.}
  \label{fig:Block-LME3}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments

gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
%\documentclass[smallextended]{svjour3}       % onecolumn (second format)
\documentclass[twocolumn]{svjour3}          % twocolumn
%
% flush right qed marks, e.g. at end of proof
\smartqed

@@

\title{Enhanced Material Point Method to face
  dynamic problems:\thanks{Funding: This
    study was funded by Agustín de Betancourt Foundation (grant number
    262390106114).}
}
\subtitle{Local-maximum entropy approximation and explicit
  predictor-corrector scheme}

@@

\begin{algorithm}
  \label{alg:PCE-algorithm}
  \floatname{algorithm}{Algorithm}
  \renewcommand{\thealgorithm}{}
  \caption{Explicit Predictor-Corrector scheme}
  \begin{algorithmic}[1]
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% º
    \STATE \textbf{Explicit Newmark Predictor} ($\gamma = 0.5$):\\
    Only in those active nodes:
    \begin{equation*}
      \vec{v}_I^{k+1} = \vec{v}_I^n + (1 - \gamma)\ \Delta t\ \vec{a}_I^k
      \quad , \quad \vec{a}_I^k = \frac{\vec{f}_{I}^{k}}{\tens{m}_I^k}
    \end{equation*}
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Impose essential boundary conditions}:\\
    At the fixed boundary, set $\vec{v}_{I}^{k+1} = 0$. 
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    % \STATE \textbf{Discard the previous nodal values}.
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Update mass matrix}:
    \begin{equation*}
      \tens{m}_{I} = \sum_{p}^{N_p}N_{Ip}^{k+1}\ m_p,
    \end{equation*}
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Deformation tensor increment calculation}.
    \begin{align*}
      &\dot{\tens{\varepsilon}_{p}}^{k+1} = \frac{1}{2}\ \sum_{I}^{N_n}
        \Grad{N_{Ip}^{k+1}}\vec{v}_{I}^{k+1} +
        (\Grad{N_{Ip}^{k+1}}\vec{v}_{I}^{k+1})^T \\
      &\Delta \tens{\varepsilon}_{p}^{k+1} = \Delta t\ \dot{\tens{\varepsilon}_{p}}^{k+1}
    \end{align*}
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Update the density field}:
    \begin{equation*}
      \rho_p^{k+1} = \frac{\rho_p^k}{1 + \mathit{tra}\left[\Delta\tens{\varepsilon}_{p}^{k+1}\right]}.
    \end{equation*}
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Balance of forces calculation}:\\
    Calculate the total grid nodal force $\vec{f}_{I}^{k+1} =
    \vec{f}_{I}^{int,k+1} + \vec{f}_{I}^{ext,k+1}$ evaluating
    \eqref{eq:nodal_internal_forces} and \eqref{eq:nodal_external_forces} in the time step $k+1$. In the
    grid node $I$ is fixed in one of the spatial dimensions, set it to
    zero to make the grid nodal acceleration zero in that direction.\\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Explicit Newmark Corrector} ($\gamma = 0.5$):
    \begin{equation*}
      \vec{v}_{I}^{k+1} = \vec{v}_{I}^{k+1} + \gamma\ \Delta t\
      \vec{a}_I^{k+1} \quad , \quad  \vec{a}_I^{k+1} = \frac{\vec{f}_{I}^{k+1}}{\tens{m}_I^{k+1}}  
    \end{equation*}
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \STATE \textbf{Update particles position and velocity}:
    \begin{align*}
      &\vec{x}_p^{k+1} = \vec{x}_p^n + \Delta t\
        \sum^{N_n}_{I} N_{Ip}^k\ \vec{v}_{I}^{k} +
        \frac{1}{2}\Delta t^2\ \sum^{N_n}_{I}\frac{N_{Ip}^k\
        \vec{f}_{I}^{k}}{\tens{m}_I^k}\\
      &\vec{v}_p^{k+1} = \vec{v}_p^n + \Delta t\
        \sum^{N_n}_{I}\frac{N_{Ip}^k\
        \vec{f}_{I}^{k}}{\tens{m}_I^k}
    \end{align*}
  \end{algorithmic}
\end{algorithm}

@@

% The previous lines describe the main algorithm without defining an
% interpolation technique, so we are free to apply a wide range of the
% existing.

The popularity of the MPM has increase notoriously during
the recent years due to its ability to deal with large strain problems
without mesh distorsion issues inherent to mesh based methods like
FE, see Zdzislaw \cite{Wieckowski2004}. However, in the simulations made with the original MPM, there are
numerical noises when particles crossing the cell boundaries. Recently
the local maximum-entropy (or local \textit{max-ent}) shape function
first introduced by Arroyo \& Ortiz (2006)\cite{Arroyo2006} has been
tested under the MPM framework by Wobbes {\it et al.} (2020)\cite{Wobbes2020} where they prof that
simulations performed with the maxent basis functions show
considerably more accurate stress approximations for MPM. Although, in
\cite{Wobbes2020} authors does not deep in how $\beta$ affects to the
accuracy and stability of the solution.

The basic idea of the shape functions based on such
an estimate is to interpret the shape function $N_I(\vec{x})$ as the
probability of $\vec{x}$ to obtain the value $\vec{x}_I$,  $I=1,
\dots, n$. Here $n$ is the number of nodes in the domain.

Taking the definition of entropy as a measure of how uncertainty a
random variable is averaged on all its possible outcomes.

Therefore following the principle of \textit{maximum-entropy}
stated by Jaynes \cite{Jaynes1957}, the least biased shape functions
are those that maximizes the entropy of the scheme

estimate is defined as the type of statistical
inference, which is the least biased possible on the given information.

Taking Shannon's entropy as the starting point:
\begin{equation}
  \label{eq:Shannon-entropy}
  H(p_1(\vec{x}),\ldots,p_n(\vec{x})) = -\sum^{N_n}_{I=1}{p_I(\vec{x})\log p_I }
\end{equation}
where $p_I(\vec{x})$ is the probability, equivalent to the mentioned
shape function $N_I(\vec{x})$,  satisfying both the zeroth and
first-order consistency.

The least-biased approximation scheme is given by
\begin{align*}
  \text{(LME)} \hspace{0.15cm} \text{Maximize} \hspace{0.15cm} &H(p) = -\sum_{I}^{N_n}{p_I(\vec{x})\log p_I } \nonumber \\
  \text{subject to} \hspace{0.15cm} &p_I \ge 0, \hspace{0.15cm} \text{I=1, ..., n} \nonumber \\
  &\sum_{I=1}^{N_n}{p_I} = 1 \nonumber \\
  &\sum_{I=1}^{N_n}{p_I \vec{x}_I} = \vec{x} \nonumber 
\end{align*}

The local max-ent approximation schemes (LME) as a Pareto set,
defined by \cite{Arroyo2006} is as follows 
\begin{align*}
  \text{(LME)}_{\beta} \hspace{0.15cm} \text{For fixed} \hspace{0.15cm}
  \vec{x} \hspace{0.15cm} \text{minimise}  \nonumber \hspace{0.15cm} &f_{\beta}(\vec{x}, p) = \beta H(\vec{x},p) - H(p) \\
  \text{subject to} \hspace{0.15cm} &p_I \ge 0, \hspace{0.2cm} \text{I=1, ..., n}  \\
                                                                   &\sum_{I=1}^{N_n}{p_I} = 1 \\
                                                                   &\sum_{I=1}^{N_n}{p_I \, \vec{x}_I} = \vec{x}
\end{align*}
for $\beta \in (0,\infty)$ is Pareto optimal. The unique solution of
the local max-ent problem (LME)$_\beta$ is: 
\begin{equation}
  \label{eq:LME-p}
p(\vec{x})=\frac{\exp\left[ -\beta \; |\vec{x}-\vec{x}_I|^2 +  \vec{\lambda}  \;  (\vec{x}-\vec{x}_I)         \right] } {Z(\vec{x},\vec{\lambda}^*(\vec{x}))}
\end{equation}
where
\begin{equation}
  \label{eq:LME-Z}
Z(\vec{x}, {\vec{\lambda}}) = \sum_{I=1}^{N_n}{ \exp \left[ -\beta \, | \vec{x}-\vec{x}_I|^2 + \vec{\lambda}  \,  (\vec{x}-\vec{x}_I)  \right]}
\end{equation}
being $\vec{\lambda}^*(\vec{x})$ the unique minimiser  for $\log Z(\vec{x}, \vec{\lambda})$

In order to obtain the first derivatives of the shape function, it is  also necessary to compute~$\nabla p^*_I$ 
\begin{equation}
  \label{eq:LME-grad-p}
\nabla p^*_I=p^*_I  \, \left(\nabla f^*_I-\sum_J^{N_n} p^*_J \, \nabla f^*_J\right)
\end{equation}
where
\begin{equation}
  \label{eq:LME-f}
f^*_I(\vec{x},  \vec{\lambda},\beta)=-\beta \, |\vec{x}-\vec{x}_I|^2 + \vec{\lambda}   \,  (\vec{x}-\vec{x}_I)
\end{equation}
Employing the chain rule, rearranging and considering $\beta$ as a constant, Arroyo and Ortiz~\cite{Arroyo2006} obtained the following expression:
\begin{eqnarray}
\nabla p^*_I &=& -p^*_I \,  (\tens{J}^*)^{-1} \,  (\vec{x} - \vec{x}_I) \label{eq26} 
\end{eqnarray}
%&&\mbox{
where \tens{J} is the Hessian matrix, defined by:
%} \nonumber \\&& \nonumber \\
\begin{eqnarray}
\tens{J}(\vec{x},  \vec{\lambda},\beta) &=& \frac{\partial \vec{r}}{\partial  \vec{\lambda}} \label{eq27} \\ 
%&& \nonumber \\
\vec{r}(\vec{x},\vec{\lambda},\beta) &\equiv& \partial_{\vec{\lambda}}{\log{ Z( \vec{x},\vec{\lambda}})} 
%\nonumber \\ &=&
  =\sum_I^{N_n} p_I(\vec{x},\vec{\lambda},\beta) \, (\vec{x} - \vec{x}_I) \label{eq28} 
\end{eqnarray}
Note that, the objective of the above procedure is to find the
$\vec{\lambda}$ which minimises  $\log Z(\vec{x}, \vec{\lambda})$. The
traditional way to obtain such a minimiser  is using Eq.~(\ref{eq27})
to calculate small increments of $\partial\vec{\lambda}$ in a
Newton-Raphson approach.

Similar to alternative non-polynomial meshfree basis functions, the
LME approximation scheme requires more than $d+1$ nodes to determine
the values of the shape functions as well as their derivatives at any
point in the convex hull of the nodal set, where $d$ is the dimension
of the problem. Due to the FE-compatibility, the LME shape function is
degenerated to linear finite element shape function if $d+1$
neighbouring nodes are chosen as the support. The support size of LME
shape functions may be controlled by adjusting a dimensionless
parameter, $\gamma=\beta h^2$ (e.g. two dimensional example shown in
figure. \ref{fig:LME_MPM})\cite{Arroyo2006}.

@@

Since  $p_I$  is defined in the entire domain, in practice, the
function $\exp(-\beta \vec{r} )$ truncated  by  a given tolerance,
10$^{-6}$, for example,  would ensure a reasonable range of
neighbours, see \cite{Arroyo2006} for details. This tolerance defines
the limit values of the influence radius and is used thereafter to
find the neighbour nodes of a given integration point. Regarding this
aspect an important note in the implementation of this interpolation
technique in the MPM is

In this research and in \cite{Arroyo2006}, $\Beta$ is a scalar as the
influence area of the shape function is controlled by the Euclidean
norm, therefore the search area is geometrically a circle in 2D, or a
sphere in 3D. Building upon the idea of anisotropic shape functions,
\cite{Kochmann2019} introduced an enhanced version of the original
local max-ent scheme, which uses an anisotropic support to deal with
tensile inestability. Nonetheless this is out of the scope of the
present document but will be incorporated in future research.

@@

\begin{figure}\sidecaption
  \centering
  \resizebox{\hsize}{!}{
    \includegraphics[width=\textwidth]{./Figures/Velocity_FE_vs_PCE_CFL_05}
  }
  \caption{Velocity evolution at the point in the bar left side.}
  \label{fig:Dyka-PCE-FE}
\end{figure}

\begin{figure}\sidecaption
  \centering
  \resizebox{\hsize}{!}{
    \includegraphics[width=\textwidth]{./Figures/Velocity_LME_gamma_comparative}
  }
  \caption{Velocity evolution at the point in the bar left side.}
  \label{fig:Dyka-LME-gamma}
\end{figure}

\begin{figure}\sidecaption
  \centering
  \resizebox{\hsize}{!}{
    \includegraphics[width=\textwidth]{./Figures/Velocity_MPM_vs_OTM_Dyka}
  }
  \caption{Velocity evolution at the point in the bar left side.}
  \label{fig:Dyka-OTM-MPM}
\end{figure}

For the convergence analysis, the root-mean-square (RMS) error in the
velocity field is computed. RMS error is defined as follows

\begin{equation}
  \label{eq:RMS}
  RMS = \sqrt{\frac{1}{N} \sum^{N}_p \left( \vect{v}_p - \hat{\vect{v}}_p \right)^2},
\end{equation}

where $\vect{v}_p$ and $\hat{\vect{v}}_p$ are respectively the analytical and
numerical solutions evaluated in the final time step in the position
of each particle.

\begin{figure*}\sidecaption
  \centering
  \resizebox{\hsize}{!}{
    \includegraphics[width=\textwidth]{./Figures/Error_evol}
  }
  \caption{Velocity evolution at the point in the bar left side.}
  \label{fig:Dyka-error-evol}
\end{figure*}

@@

\begin{figure}
  \centering
  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
  \subfloat[t = 0 seconds.]{
    \includegraphics[width=0.8\columnwidth]{Figures/Block_LME3_PCE_a_t0}
    \label{fig:Block-LME3-PCE-t0}
  }
  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
  \qquad
  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
  \subfloat[t = 5 seconds.]{
    \includegraphics[width=0.8\columnwidth]{Figures/Block_LME3_PCE_b_t025}
    \label{fig:Block-LME3-PCE-t025}
  }
  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \qquad
  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
  \subfloat[t = 20 seconds]{
    \includegraphics[width=0.8\columnwidth]{Figures/Block_LME3_PCE_c_t1}
    \label{fig:Block-LME3-PCE-t1}
  }
  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
  \caption{Vertical normal stress and position of material points
    during the loading process for a soft soil ($E = 5\ MPa$, $\rho_0
    = 6\cdot 10^3\ kg/m^3$). Numerical parameters considered for the
    simulation are : Local \textit{max-ent} shape function $\gamma =3$
    and explicit PC scheme with CFL 0.1.}
  \label{fig:Block-LME3}
\end{figure}
